---
title: "Fair LIHTC EDA"
author: "Colin Ries"
header-includes:
- \usepackage{booktabs}
- \usepackage{siunitx}
- \newcolumntype{d}{S[input-symbols = ()]}
output:
  pdf_document: default
  html_notebook: default
---

```{r setup, include=FALSE}

knitr::opts_chunk$set(warning=FALSE, results = FALSE)

```

```{r packages, include=FALSE}

library(tidyverse)
library(haven)
library(readxl)
library(estimatr)
library(modelsummary)

```

# Data

## LIHTC Property-Level Data

Downloaded LIHTC property data from [https://lihtc.huduser.gov/lihtcpub.zip](https://lihtc.huduser.gov/lihtcpub.zip) on April 22, 2022.

```{r tidy_lihtc_data}
# Load National LIHTC data
lihtc <- read_csv('../data/LIHTCPUB.CSV')

# Normalize the dummy variables
lihtc <- lihtc %>% 
  mutate(qct = 0 + (qct==1)) %>% 
  # Do not distinguish between metro, non-metro, and gulf opportunity zone DDAs
  mutate(dda = 0 + (dda==1)) %>% 
  mutate(metro = 0 + (metro!=3)) %>% 
  mutate(non_prof = 0 + (non_prof==1)) %>% 
  mutate(trgt_fam = 0 + (trgt_fam==1)) %>% 
  mutate(trgt_hml = 0 + (trgt_hml==1)) %>% 
  mutate(trgt_eld = 0 + (trgt_eld==1)) %>% 
  mutate(nine_prcnt = 0 + (credit==1))

```

```{r test_lihtc_data}

# Allocation data for new developments in metro areas 
lihtc <- lihtc %>% 
  filter(yr_alloc <= 2015) %>% 
  filter(type == 1) %>% 
  filter(metro == 1)

# Calculate the number of projects allocated credits by year
ggplot(data = lihtc, mapping = aes(x=yr_alloc)) +
  geom_freqpoly(binwidth=1) + xlim(1987,2015)

# Calculate the total number of low-income units for each allocation year
li_units <- lihtc %>% 
  group_by(yr_alloc) %>% 
  summarise(y = sum(li_units, na.rm=TRUE))

ggplot(data = li_units, mapping = aes(x=yr_alloc, y=y)) +
  geom_line() + xlim(1987,2015)

```

# Concentrations of Poverty

The primary critique of the LIHTC program is that it incentivizes perpetuating concentrations of poverty by granting a basis boost for developments in QCTs. We examine whether siting of developments in QCTs has increased over time. 

```{r qct_model}

lihtc %>%
  group_by(yr_alloc) %>%
  summarize(prop = mean(qct, na.rm=TRUE)) %>%
  ggplot(aes(x=yr_alloc,y=prop)) +
    geom_line()

qct_model <-lm_robust(qct~yr_alloc+non_prof+trgt_pop+trgt_fam+trgt_eld+trgt_hml+nine_prcnt, 
                      data = lihtc, fixed_effects = proj_st)

modelsummary(qct_model, stars = TRUE, output = 'latex')

```
# Difficult to Develop Areas

```{r dda_model}

lihtc %>%
  group_by(yr_alloc) %>%
  summarize(prop = mean(dda, na.rm=TRUE)) %>%
  ggplot(aes(x=yr_alloc,y=prop)) +
    geom_line()

dda_model <-lm_robust(dda~yr_alloc+non_prof+trgt_pop+trgt_fam+trgt_eld+trgt_hml+nine_prcnt, 
                      data = lihtc, fixed_effects = proj_st)

modelsummary(dda_model, stars = TRUE, output = 'latex')

```

# Retroactively Examining SADDAs

Downloaded Small Area DDA data from [https://www.huduser.gov/portal/datasets/fmr/fmr2022/fy2022_safmrs.xlsx](https://www.huduser.gov/portal/datasets/fmr/fmr2022/fy2022_safmrs.xlsx) on April 28, 2022.

Downlaoded the 2010 Decennial ZCTA to FIPS crosswalk from [https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.2010.html](https://www.census.gov/geographies/reference-files/time-series/geo/relationship-files.2010.html) on May 7th, 2022.

```{r tidy_saddas_data}

# Import 2022 Metropolitan DDAs
sadda <- read_xlsx('../data/2022-DDAs-Data-Used-to-Designate.xlsx', sheet=1)

# Import zcta to FIPS crosswalk
zcta_to_ct <- read_csv("../data/zcta_tract_rel_10.txt")

# Identify 2022 Small Area DDA by ZCTA
sadda <- sadda %>% 
  rename(sdda = "2022 SDDA (1=SDDA)") %>%
  rename(zcta = "ZIP Code Tabulation Area (ZCTA)") %>%
  filter(sdda == 1) 

# Identify all Census Tracts in 2022 SADDAs
zcta_to_ct <-  zcta_to_ct %>% 
  rename(zcta = ZCTA5) %>%
  rename(fips = GEOID) %>% 
  filter(zcta %in% sadda$zcta)

lihtc <- lihtc %>% 
  mutate(sadda = 0 + (fips2010 %in% zcta_to_ct$fips))

# Import Non-Metropolitan DDAs
# nm_dda <- read_xlsx('../data/2022-DDAs-Data-Used-to-Designate.xlsx', sheet=2)

# nm_dda <- nm_dda %>% 
#   rename(nmdda = '2022 NMDDA') %>% 
#   rename(cnty_fips = FIPS)
#   filter(nmdda ==0)

```

```{r sadda_model}

lihtc %>%
  group_by(yr_alloc) %>%
  summarize(prop = mean(sadda, na.rm=TRUE)) %>%
  ggplot(aes(x=yr_alloc,y=prop)) +
    geom_line()

sadda_model <-lm_robust(sadda~qct+yr_alloc+non_prof+trgt_pop+trgt_fam+trgt_eld+trgt_hml+nine_prcnt, 
                        data = lihtc, fixed_effects = proj_st)

modelsummary(sadda_model, stars = TRUE, output = 'latex')

```

```{r op_nbh}

lihtc <- lihtc %>% 
  mutate(sadda = 0 + (fips2010 %in% zcta_to_ct$fips)) %>% 
  mutate(op_nbh = 0 + (sadda & !qct)) 
  
lihtc %>%
  group_by(yr_alloc) %>%
  summarize(prop = mean(op_nbh, na.rm=TRUE)) %>%
  ggplot(aes(x=yr_alloc,y=prop)) +
    geom_line()

```






```{r op_nbh_model}

op_nbh_model <-lm_robust(op_nbh~yr_alloc+non_prof+trgt_pop+trgt_fam+trgt_eld+trgt_hml+nine_prcnt, 
                         data = lihtc, fixed_effects = proj_st)

modelsummary(op_nbh_model, stars = TRUE, output = 'figures/op_nbh.png')

```

